---
title: "RestaurantDataAnalysis"
output:
  html_document: default
  pdf_document: default
date: "2025-06-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}
# Load required libraries

library(utils)  # for download.file and unzip
library(readr)  
library(tidyverse)
library(stringr)
library(lubridate)
```

# Download data 

```{r}
# Create lists of csv files 
csv_list_all <- list.files("./All_csvs")

csv_list_fs <- csv_list_all[grepl("722511", csv_list_all)] # Full service restaurants
csv_list_ls <- csv_list_all[grepl("722513", csv_list_all)] # Limited service restaurants
csv_list_b <- csv_list_all[grepl("7224", csv_list_all)] # Drinking places 

# Create empty data.frame

cols <- names(read_csv(file = paste0("./All_csvs/", csv_list_all[1]))) # Get list of column names 
df = data.frame(matrix(nrow = 0, ncol = length(cols))) # Initialize empty df
colnames(df) = cols


# Create vector of variables of interest

vars_list <- c("area_fips", "area_title", "own_code", "industry_code", "agglvl_code", "year", "qtr", "qtrly_estabs_count", "month1_emplvl", "month2_emplvl", "month3_emplvl", "avg_wkly_wage")

fips_list <- c("11001", "24000", "51000")

```


```{r}
# Loop through csv lists to load all csvs by NAICS code
      ## vars_list, fips_list defined above

df_list_fs <- lapply(csv_list_fs, function(x) 
  filter(select(read_csv(file = paste0("./All_csvs/", x)), vars_list), area_fips %in% fips_list))

df_list_ls <- lapply(csv_list_ls, function(x) 
  filter(select(read_csv(file = paste0("./All_csvs/", x)), vars_list), area_fips %in% fips_list))

df_list_b <- lapply(csv_list_b, function(x) 
  filter(select(read_csv(file = paste0("./All_csvs/", x)), vars_list), area_fips %in% fips_list))
```


```{r}
# Rbind tables  

df_fs <- do.call(rbind, df_list_fs)

df_ls <- do.call(rbind, df_list_ls)

df_b <- do.call(rbind, df_list_b)


```

## Clean data tables

```{r}

# Convert quarter & year variables to dates 
## Creating "date" with the structure "%Y-%m-%d" for the start of each qtr

df_fs <- df_fs %>% mutate(month = case_when(qtr==1 ~ "01",
                                   qtr==2 ~ "04",
                                   qtr==3 ~ "07",
                                   qtr==4 ~ "10"),
                   date = as.Date(paste(year, month, "01", sep = "-")))



df_ls <- df_ls %>% mutate(month = case_when(qtr==1 ~ "01",
                                   qtr==2 ~ "04",
                                   qtr==3 ~ "07",
                                   qtr==4 ~ "10"),
                   date = as.Date(paste(year, month, "01", sep = "-")))


df_b <- df_b %>% mutate(month = case_when(qtr==1 ~ "01",
                                   qtr==2 ~ "04",
                                   qtr==3 ~ "07",
                                   qtr==4 ~ "10"),
                   date = as.Date(paste(year, month, "01", sep = "-")))

```


```{r}
# Restructure datasets
## Data for each NAICS code is further categorize by "own_code"
## Own code 1 == Federal government 
## Own code 5 == Private
## The data should be restructured to show the sum across all ownership types

df_fs_sums <- df_fs %>% group_by(area_fips, area_title, industry_code, year, qtr, date) %>% 
  summarise(all_qtrly_estabs_count = sum(qtrly_estabs_count),
            all_month1_emplvl = sum(month1_emplvl),
            all_month2_emplvl = sum(month2_emplvl),
            all_month3_emplvl = sum(month3_emplvl)) %>% ungroup()

df_ls_sums <- df_ls %>% group_by(area_fips, area_title, industry_code, year, qtr, date) %>% 
  summarise(all_qtrly_estabs_count = sum(qtrly_estabs_count),
            all_month1_emplvl = sum(month1_emplvl),
            all_month2_emplvl = sum(month2_emplvl),
            all_month3_emplvl = sum(month3_emplvl)) %>% ungroup()


df_b_sums <- df_b %>% group_by(area_fips, area_title, industry_code, year, qtr, date) %>% 
  summarise(all_qtrly_estabs_count = sum(qtrly_estabs_count),
            all_month1_emplvl = sum(month1_emplvl),
            all_month2_emplvl = sum(month2_emplvl),
            all_month3_emplvl = sum(month3_emplvl)) %>% ungroup()

```


```{r}
df_b_sums %>% select(year) %>% unique()
```


# Exploring the data 

## Plot 1: Growth of full-service establishments in DC (Index==2018)

```{r}
## Subset for plot

index_year <- 2018


p1_subset <- df_fs_sums %>% mutate(type = "Full-Service") 
  ##%>% filter(year >= index_year)

# Indexes for [Year]
dc_index <- p1_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==11001) %>% select(all_qtrly_estabs_count) %>% pull()
md_index <- p1_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==24000) %>% select(all_qtrly_estabs_count) %>% pull()
va_index <- p1_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==51000) %>% select(all_qtrly_estabs_count) %>% pull()

p1_subset <- p1_subset %>% mutate(indexed_qtrly_estabs_count = round(case_when(area_fips==11001 ~ 100*(all_qtrly_estabs_count/dc_index),
                                                       area_fips==24000 ~ 100*(all_qtrly_estabs_count/md_index),
                                                       area_fips==51000 ~ 100*(all_qtrly_estabs_count/va_index)),2))

p1_subset$labels <- NA
p1_subset$labels[which(p1_subset$date == max(p1_subset$date))] <- p1_subset$indexed_qtrly_estabs_count[which(p1_subset$date == max(p1_subset$date))]

```

```{r warning=F}
## Plot

  # Range from 80-160 
  # Colors -- DC = blue, VA = green, MD = orange

p1 <- ggplot(p1_subset, aes(x = date, y = indexed_qtrly_estabs_count, color = as.factor(area_fips))) +
  geom_line(linewidth = 1) +
  theme_classic() +
  geom_text(aes(label=p1_subset$labels, color = as.factor(area_fips)), show.legend=FALSE, hjust = 0) +
  #geom_label(label = p1_subset$labels, nudge_x = 1, warning=F, show.legend = FALSE) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y/%m", expand=c(0,0), limits = as.Date(c(paste0("2018", "-01-01"), "2025-01-01")))


p1 + 
  theme(legend.position = c(0.2,0.8),
        axis.text.x = element_text(angle = 45, vjust=0.5),
        plot.title = element_text(margin=margin(0,0,15,0)),
        plot.subtitle = element_text(size=9)) +
  scale_y_continuous(limits = c(80,140)) +
  scale_color_manual(labels = c("District of Columbia", "Maryland", "Virginia"), name = "State", 
                     values=c("darkblue", "darkorange", "darkgreen")) +
  labs(title = paste0("Growth in number of full-service establishments in DC, MD, and VA since ", index_year),
       subtitle = paste0("Index: ",index_year,"Q1=100"),
       x = "",
       y = "",
       caption = paste0("An increase of 1 represents a 1% increase above the ",index_year," restaurant count for that state")) 


```

## Plot 2: Growth of full-service establishments in DC (Index==2021)

```{r}
## Subset for plot

index_year <- 2021


p2_subset <- df_fs_sums %>% mutate(type = "Full-Service") 
  ##%>% filter(year >= index_year)

# Indexes for [Year]
dc_index <- p2_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==11001) %>% select(all_qtrly_estabs_count) %>% pull()
md_index <- p2_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==24000) %>% select(all_qtrly_estabs_count) %>% pull()
va_index <- p2_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==51000) %>% select(all_qtrly_estabs_count) %>% pull()

p2_subset <- p2_subset %>% mutate(indexed_qtrly_estabs_count = round(case_when(area_fips==11001 ~ 100*(all_qtrly_estabs_count/dc_index),
                                                       area_fips==24000 ~ 100*(all_qtrly_estabs_count/md_index),
                                                       area_fips==51000 ~ 100*(all_qtrly_estabs_count/va_index)),2))

p2_subset$labels <- NA
p2_subset$labels[which(p2_subset$date == max(p2_subset$date))] <- p2_subset$indexed_qtrly_estabs_count[which(p2_subset$date == max(p2_subset$date))]

```

```{r warning=F}
## Plot

  # Range from 80-160 
  # Colors -- DC = blue, VA = green, MD = orange

p2 <- ggplot(p2_subset, aes(x = date, y = indexed_qtrly_estabs_count, color = as.factor(area_fips))) +
  geom_line(linewidth = 1) +
  theme_classic() +
  geom_text(aes(label=p2_subset$labels, color = as.factor(area_fips)), show.legend=FALSE, hjust = 0) +
  #geom_label(label = p2_subset$labels, nudge_x = 1, warning=F, show.legend = FALSE) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y/%m", expand=c(0,0), limits = as.Date(c(paste0("2018", "-01-01"), "2025-01-01")))


p2 + 
  theme(legend.position = c(0.2,0.8),
        axis.text.x = element_text(angle = 45, vjust=0.5),
        plot.title = element_text(margin=margin(0,0,15,0)),
        plot.subtitle = element_text(size=9)) +
  scale_y_continuous(limits = c(80,140)) +
  scale_color_manual(labels = c("District of Columbia", "Maryland", "Virginia"), name = "State", 
                     values=c("darkblue", "darkorange", "darkgreen")) +
  labs(title = paste0("Growth in number of full-service establishments in DC, MD, and VA since ", index_year),
         subtitle = paste0("Index: ",index_year,"Q1=100"),
       x = "",
       y = "",
       caption = paste0("An increase of 1 represents a 1% increase above the ",index_year," restaurant count for that state")) 

```


## Plot 3: Growth of full-service establishments in DC (Index==2023)

```{r}
## Subset for plot

index_year <- 2023


p3_subset <- df_fs_sums %>% mutate(type = "Full-Service") 
  ##%>% filter(year >= index_year)

# Indexes for [Year]
dc_index <- p3_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==11001) %>% select(all_qtrly_estabs_count) %>% pull()
md_index <- p3_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==24000) %>% select(all_qtrly_estabs_count) %>% pull()
va_index <- p3_subset %>% filter(date==paste0(index_year,"-01-01"), area_fips==51000) %>% select(all_qtrly_estabs_count) %>% pull()

p3_subset <- p3_subset %>% mutate(indexed_qtrly_estabs_count = round(case_when(area_fips==11001 ~ 100*(all_qtrly_estabs_count/dc_index),
                                                       area_fips==24000 ~ 100*(all_qtrly_estabs_count/md_index),
                                                       area_fips==51000 ~ 100*(all_qtrly_estabs_count/va_index)),2))

p3_subset$labels <- NA
p3_subset$labels[which(p3_subset$date == max(p3_subset$date))] <- p3_subset$indexed_qtrly_estabs_count[which(p3_subset$date == max(p3_subset$date))]

```

```{r warning=F}
## Plot

  # Range from 80-160 
  # Colors -- DC = blue, VA = green, MD = orange

p3 <- ggplot(p3_subset, aes(x = date, y = indexed_qtrly_estabs_count, color = as.factor(area_fips))) +
  geom_line(linewidth = 1) +
  theme_classic() +
  geom_text(aes(label=p3_subset$labels, color = as.factor(area_fips)), show.legend=FALSE, hjust = 0) +
  #geom_label(label = p3_subset$labels, nudge_x = 1, warning=F, show.legend = FALSE) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y/%m", expand=c(0,0), limits = as.Date(c(paste0("2018", "-01-01"), "2025-01-01")))


p3 + 
  theme(legend.position = c(0.2,0.8),
        axis.text.x = element_text(angle = 45, vjust=0.5),
        plot.title = element_text(margin=margin(0,0,15,0)),
        plot.subtitle = element_text(size=9)) +
  scale_y_continuous(limits = c(80,140)) +
  scale_color_manual(labels = c("District of Columbia", "Maryland", "Virginia"), name = "State", 
                     values=c("darkblue", "darkorange", "darkgreen")) +
  labs(title = paste0("Growth in number of full-service establishments in DC, MD, and VA since ", index_year),
         subtitle = paste0("Index: ",index_year,"Q1=100"),
       x = "",
       y = "",
       caption = paste0("An increase of 1 represents a 1% increase above the ",index_year," restaurant count for that state")) 


```

## Plot 4: Full and Limited service employment levels in DC since Covid 

```{r}

# Subset and restructure data 
## Employment level data is available by month 
## Pivot month-specific columns into one column

## Full-service establishments

p4_fs <- df_fs_sums %>% pivot_longer(cols = c("all_month1_emplvl", "all_month2_emplvl", "all_month3_emplvl"),
                                    names_to = "Qtr_month", values_to = "employment_lvl") %>% 
           mutate(type = "Full-service restaurants",
                  month = case_when(qtr==1 & Qtr_month=="all_month1_emplvl" ~ 01,
                           qtr==1 & Qtr_month=="all_month2_emplvl" ~ 02,
                           qtr==1 & Qtr_month=="all_month3_emplvl" ~ 03,
                           qtr==2 & Qtr_month=="all_month1_emplvl" ~ 04,
                           qtr==2 & Qtr_month=="all_month2_emplvl" ~ 05,
                           qtr==2 & Qtr_month=="all_month3_emplvl" ~ 06,
                           qtr==3 & Qtr_month=="all_month1_emplvl" ~ 07,
                           qtr==3 & Qtr_month=="all_month2_emplvl" ~ 08,
                           qtr==3 & Qtr_month=="all_month3_emplvl" ~ 09,
                           qtr==4 & Qtr_month=="all_month1_emplvl" ~ 10,
                           qtr==4 & Qtr_month=="all_month2_emplvl" ~ 11,
                           qtr==4 & Qtr_month=="all_month3_emplvl" ~ 12),
                   date = as.Date(paste(year, month, "01", sep="-"))) 


## Limited-service establishments

p4_ls <- df_ls_sums %>% pivot_longer(cols = c("all_month1_emplvl", "all_month2_emplvl", "all_month3_emplvl"),
                                    names_to = "Qtr_month", values_to = "employment_lvl") %>% 
           mutate(type = "Limited-service restaurants",
                  month = case_when(qtr==1 & Qtr_month=="all_month1_emplvl" ~ 01,
                           qtr==1 & Qtr_month=="all_month2_emplvl" ~ 02,
                           qtr==1 & Qtr_month=="all_month3_emplvl" ~ 03,
                           qtr==2 & Qtr_month=="all_month1_emplvl" ~ 04,
                           qtr==2 & Qtr_month=="all_month2_emplvl" ~ 05,
                           qtr==2 & Qtr_month=="all_month3_emplvl" ~ 06,
                           qtr==3 & Qtr_month=="all_month1_emplvl" ~ 07,
                           qtr==3 & Qtr_month=="all_month2_emplvl" ~ 08,
                           qtr==3 & Qtr_month=="all_month3_emplvl" ~ 09,
                           qtr==4 & Qtr_month=="all_month1_emplvl" ~ 10,
                           qtr==4 & Qtr_month=="all_month2_emplvl" ~ 11,
                           qtr==4 & Qtr_month=="all_month3_emplvl" ~ 12),
                   date = as.Date(paste(year, month, "01", sep="-"))) 


## Bars and drinking places

p4_b <- df_b_sums %>% pivot_longer(cols = c("all_month1_emplvl", "all_month2_emplvl", "all_month3_emplvl"),
                                    names_to = "Qtr_month", values_to = "employment_lvl") %>% 
           mutate(type = "Bars and drinking places",
                  month = case_when(qtr==1 & Qtr_month=="all_month1_emplvl" ~ 01,
                           qtr==1 & Qtr_month=="all_month2_emplvl" ~ 02,
                           qtr==1 & Qtr_month=="all_month3_emplvl" ~ 03,
                           qtr==2 & Qtr_month=="all_month1_emplvl" ~ 04,
                           qtr==2 & Qtr_month=="all_month2_emplvl" ~ 05,
                           qtr==2 & Qtr_month=="all_month3_emplvl" ~ 06,
                           qtr==3 & Qtr_month=="all_month1_emplvl" ~ 07,
                           qtr==3 & Qtr_month=="all_month2_emplvl" ~ 08,
                           qtr==3 & Qtr_month=="all_month3_emplvl" ~ 09,
                           qtr==4 & Qtr_month=="all_month1_emplvl" ~ 10,
                           qtr==4 & Qtr_month=="all_month2_emplvl" ~ 11,
                           qtr==4 & Qtr_month=="all_month3_emplvl" ~ 12),
                   date = as.Date(paste(year, month, "01", sep="-"))) 


p4_subset <- rbind(p4_fs, p4_ls, p4_b)

p4_subset_dc <- p4_subset %>% filter(area_fips=="11001")

```


```{r}
# Plot 

## Data is p4_subset or p4_subset_dc created above

p4 <- ggplot(p4_subset_dc, aes(x = date, y = employment_lvl, color = type, group = type)) +
  expand_limits(y = 45000) +
  geom_line() +
  geom_vline(xintercept = as.Date("2023-05-01"), linetype="dashed", color = "black") +
  geom_vline(xintercept = as.Date("2023-07-01"), linetype="dashed", color = "black") +
  geom_vline(xintercept = as.Date("2024-07-01"), linetype="dashed", color = "black") +
  geom_text(aes(x=as.Date("2021-11-01"), 
                label="Initiative I82 Increases: \n May 2023:  $5.35 to $6.00 \n July 2023: $6:00 to $8.00 \n July 2024: $8.00 to $10.00", 
                y=35000), size = 3, colour="black") +
  theme_classic() +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y", expand = c(0,0))

p4 + 
  theme(legend.position = c(0.2,0.85)) + 
  scale_y_continuous(breaks=seq(0,45000,10000)) +
  scale_color_brewer(palette="Dark2") +
  labs(title = "Full and Limited service employment levels in DC have rebounded since Covid",
       x = "Year", y = "Number of Employees")

```

## Plot 5: Full and Limited service restaurant employment levels in DC since I82 implementation 

```{r}
p5_subset <- p4_subset_dc %>% filter(year >= 2022)

p5_subset_labels <- p5_subset
p5_subset_labels$labels <- NA

# Adding maximum value label 
p5_subset_labels$labels[which(p5_subset_labels$date == max(p5_subset_labels$date))] <- p5_subset_labels$employment_lvl[which(p5_subset_labels$date == max(p5_subset_labels$date))] 

# Adding label for May 2023
p5_subset_labels$labels[which(p5_subset_labels$date == as.Date("2023-05-01"))] <- p5_subset_labels$employment_lvl[which(p5_subset_labels$date == as.Date("2023-05-01"))] 

```

```{r, warning=FALSE}
# Plot 5

## Resource: https://r-graph-gallery.com/275-add-text-labels-with-ggplot2.html 

p5 <- ggplot(p5_subset_labels, aes(x = date, y = employment_lvl, color = type, group = type)) +
  expand_limits(x = as.Date("2024-12-01"), y = 45000) +
  geom_line(size=1) +
  geom_vline(xintercept = as.Date("2023-05-01"), linetype="dashed", color = "black") +
  geom_vline(xintercept = as.Date("2023-07-01"), linetype="dashed", color = "black") +
  geom_vline(xintercept = as.Date("2024-07-01"), linetype="dashed", color = "black") +
  geom_label(label = p5_subset_labels$labels, nudge_x = 1, warning=F, show.legend = FALSE) +
  geom_text(aes(x=as.Date("2022-07-01"), 
                label="Initiative I82 Increases: \nMay 2023:  $5.35 to $6.00 \nJuly 2023: $6:00 to $8.00 \nJuly 2024: $8.00 to $10.00", 
                y=20000, fontface = "plain"), size = 3, colour="black", hjust = 0) +
  theme_classic() +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y/%m", expand = c(0,0))

p5 + 
  theme(legend.position = c(0.2,0.85),
        axis.text.x=element_text(angle = 45, vjust=0.5),
        plot.title = element_text(margin=margin(0,0,15,0))) +
  scale_y_continuous(breaks=seq(0,45000,5000)) +
  scale_color_brewer(palette="Dark2") +
  labs(title = "Restaurant employment levels in DC since I82 implementation",
       x = "Date", y = "Number of Employees")

```

## Table 1: 





